---
title: "Gina Nichols - GAMs"
author: "Miranda Tilton"
date: "October 30, 2020"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

# Data processing

```{r warning = FALSE, message = FALSE}
# remotes::install_github("vanichols/maRsden")
library(magrittr)
library(maRsden)
library(dplyr)
library(ggplot2)
library(lemon) # chunk option `render = lemon_print` makes tables prettier
```

```{r render = lemon_print, message = FALSE}
myd <- 
  mrs_penetrom %>% 
  left_join(mrs_plotkey) %>% 
  #filter(year != "2020") %>%
  mutate(resis_Mpa = resis_kpa/1000) %>%
  select(year, doy, block, rot_trt, plot_id, rep_id, depth_cm, resis_Mpa) %>% 
  arrange(block, plot_id, rep_id, depth_cm)

# make new factor variables and convert old trt/block variables into factors
# sorry, I couldn't figure out how to do this in mutate() without gnarly warnings
myd$year_doy <- as.factor(paste(myd$year, myd$doy, sep = "_"))
myd$trt_yr <- as.factor(paste(myd$rot_trt, myd$year, myd$doy, sep = "_"))
# myd$trt_block_yr <- as.factor(paste(myd$rot_trt, myd$block, myd$year, myd$doy, sep = "_"))
myd$block <- as.factor(myd$block)
myd$rot_trt <- as.factor(myd$rot_trt)

head(myd, 12)
```

\newpage

# Data visualization

```{r fig.width = 7, fig.height = 8}
# may want to account for interaction w/ year? looks like pattern changes over time
ggplot(data = myd) + 
  geom_line(aes(x = depth_cm, y = resis_Mpa, group = rep_id, color = year_doy)) +
  facet_wrap( ~ block + rot_trt, ncol = 2)
```

\newpage

# Fit and check GAM

```{r model_fitting, warning = FALSE, message = FALSE, cache = TRUE}
library(mgcv)
mod1 <- gam(resis_Mpa ~ s(depth_cm, by = trt_yr, bs = "cr", k = 8) + trt_yr, 
           data = myd, method = "REML")
```

```{r fig.height = 5}
# plot(mod, residuals = TRUE, shade = TRUE)
par(mar = c(4, 4, 3, 0))
mgcv::gam.check(mod1)
```

To [check a GAM with `gam.check()`](https://rdrr.io/cran/mgcv/man/gam.check.html), we look for a few things:

- If `edf` is too close to `k'`, we may need more knots
- `k-index` is the estimate divided by the residual variance. The further below 1 this is, the more likely it is that there is missed pattern left in the residuals.
- The p-value for the `k-index` is computed by simulation: the residuals are randomly re-shuffled `k.rep` times to obtain the null distribution of the differencing variance estimator, if there is no pattern in the residuals.
- If the p-value is too close to zero, there is a significant pattern in the residuals that should be addressed.

In this case, the `edf` values are not too close to 7, and the `k-index` is close to 1, but the residuals are heteroskedastic (i.e., have a megaphone shape), likely because the curves are all close to zero at zero depth but spread out a lot for higher depths.

Next, I transform `resis_Mpa` with the square root (since all resistance values are non-negative) and fit another model, to unify/control variances at high depths.

\newpage

# Adjust for non-constant variance of `depth_cm` and refit

```{r model_refitting, warning = FALSE, message = FALSE, cache = TRUE}
mod2 <- gam(sqrt(resis_Mpa) ~ s(depth_cm, by = trt_yr, bs = "cr", k = 8) + trt_yr, 
           data = myd, method = "REML")
```

```{r fig.height = 5}
# plot(mod, residuals = TRUE, shade = TRUE)
par(mar = c(4, 4, 3, 0))
mgcv::gam.check(mod2)
```

Looking at the residuals plots for each group (two pages further down this document), it seems that there is still an issue with non-constant variance, but this time by year/doy instead of depth. To investigate, I remove the problem year and fit the model one more time.

\newpage

## View fitted model by group

```{r fig.width = 6, fig.height = 3}
# view the 24 fitted curves to visually inspect differences
myd$p <- predict(mod2)

ggplot(data = myd) +
  geom_line(aes(x = depth_cm, y = p^2, color = year_doy, group = year_doy)) + # note p^2
  facet_wrap( ~ rot_trt, ncol = 2) +
  guides(color = FALSE)
```

\newpage

## View residuals by group

```{r fig.width = 7, fig.height = 3.5, out.width = "80%"}
myd$resid <- mod2$residuals

ggplot(data = myd) + 
  geom_line(aes(x = depth_cm, y = resid, group = rep_id, color = year_doy)) +
  facet_wrap( ~ trt_yr, ncol = length(unique(myd$year_doy)))
```

## View data with fitted curves by group

```{r fig.width = 7, fig.height = 3.5, out.width = "80%"}
ggplot(data = myd) + 
  geom_line(aes(x = depth_cm, y = resis_Mpa, group = rep_id, color = year_doy)) +
  geom_line(aes(x = depth_cm, y = p^2, group = rep_id), color = "black") + # note p^2
  facet_wrap( ~ trt_yr, ncol = length(unique(myd$year_doy)))
```

\newpage

# Remove year that is most unlike the others and (re)refit

```{r model_rerefitting, warning = FALSE, message = FALSE, cache = TRUE}
myd2 <- filter(myd, year != 2020)
myd2$trt_yr <- as.factor(as.character(myd2$trt_yr))

mod3 <- gam(sqrt(resis_Mpa) ~ s(depth_cm, by = trt_yr, bs = "cr", k = 12) + trt_yr,
           data = myd2, method = "REML")
```

```{r fig.height = 5}
# plot(mod, residuals = TRUE, shade = TRUE)
par(mar = c(4, 4, 3, 0))
mgcv::gam.check(mod3)
```

Whoops, that made the p-values worse... Let's stick with the second model?

\newpage

## View fitted model by group

```{r fig.width = 6, fig.height = 3}
# view the 24 fitted curves to visually inspect differences
myd2$p <- predict(mod3)

ggplot(data = myd2) +
  geom_line(aes(x = depth_cm, y = p^2, color = year_doy, group = year_doy)) + # note p^2
  facet_wrap( ~ rot_trt, ncol = 2) +
  guides(color = FALSE)
```

\newpage

## View residuals by group

```{r fig.width = 7, fig.height = 3.5, out.width = "80%"}
myd2$resid <- mod3$residuals

ggplot(data = myd2) + 
  geom_line(aes(x = depth_cm, y = resid, group = rep_id, color = year_doy)) +
  facet_wrap( ~ trt_yr, nrow = 2)
```

## View data with fitted curves by group

```{r fig.width = 7, fig.height = 3.5, out.width = "80%"}
ggplot(data = myd2) + 
  geom_line(aes(x = depth_cm, y = resis_Mpa, group = rep_id, color = year_doy)) +
  geom_line(aes(x = depth_cm, y = p^2, group = rep_id), color = "black") + # note p^2
  facet_wrap( ~ trt_yr, nrow = 2)
```

\newpage

# Differences between treatments - 1st way

To test for differences between treatments, we'll use the following function from [https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/](https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/):

```{r}
smooth_diff <- function(model, newdata, f1, f2, var, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    r1 <- newdata[[var]] == f1
    r2 <- newdata[[var]] == f2
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}
```

We want to compare the differences between treatments, within years, and averaging over blocks. First, we set up a data frame that pairs the variable levels between treatments, within year/doy values (and leaving out mention of blocks).

```{r render = lemon_print}
# get all combinations year/doy (excluding block this time)
base_str <- unique(myd$year_doy)

# add trt to the front of each base str, yielding pairwise df
var_df <- data.frame(var1 = paste("2y", base_str, sep = "_"),
                     var2 = paste("4y", base_str, sep = "_"),
                     stringsAsFactors = FALSE)

head(var_df)
```

\newpage

Next, we create a data frame for prediction that contains the depth values and levels of `trt_yr`. (NOTE: using `mod2`)

```{r, render = lemon_print}
newdata <- expand.grid(depth_cm = unique(myd$depth_cm),
                        trt_yr = levels(myd$trt_yr))

out <- purrr::map_dfr(1:nrow(var_df), function(i) {
  d <- smooth_diff(mod2, newdata, 
              f1 = var_df[i,1], f2 = var_df[i,2], 
              var = "trt_yr")
  d$pair <- as.character(d$pair) # prevent map_dfr from combining factors
  return(d)
})

comp <- cbind(depth_cm = unique(myd$depth_cm), out) # add depth values
comp$pair <- as.factor(comp$pair) # make this a factor again for ggplot2
head(comp)
```

Now, we can plot the difference between treatments, with associated confidence bands, for each block and year/doy combination. Any depth values where the shading does not cross the dashed red line (at diff = 0) are points where the treatment resistances differ significantly. \textcolor{red}{Note: this is all still on the sqrt scale. In other words, this is the difference of the expected square root of the response.} To conduct inference on the original scale, check out the next section.

```{r fig.width = 6, fig.height = 4}
ggplot(comp, aes(x = depth_cm, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
    geom_hline(aes(yintercept = 0), colour="#990000", linetype="dashed") +
    facet_wrap(~ pair, ncol = 2) +
    labs(x = NULL, y = 'Difference in resis_Mpa trend') +
  ggtitle("Difference in resis_Mpa trend", subtitle = "Sqrt scale")
```

\newpage

# Differences between treatments - 2nd way

Here's a quick intro to `emmeans` by example:

```{r warning = FALSE}
library(emmeans)
em <- emmeans(mod2, specs = "trt_yr") # get estimated marginal means
em
pairs(em) # get pairwise contrasts
```

\newpage

You can specify `specs = pairwise ~ var` to get the contrasts at the same time as the estimates.

```{r}
em <- emmeans(mod2, specs = pairwise ~ "trt_yr")
em
```

\newpage

Even better, `type = "response"` will transform estimates back from the square-root scale to original scale. (Here, contrasts are still on sqrt scale, more info [here](https://cran.r-project.org/web/packages/emmeans/vignettes/comparisons.html))

```{r}
em <- emmeans(mod2, specs = pairwise ~ "trt_yr", type = "response")
em
```

\newpage

Technically, these are estimates for depth = `mean(depth_cm)`. Say we want to specify estimating for `depth_cm = 5`:

```{r}
em <- emmeans(mod2, specs = pairwise ~ "trt_yr", type = "response", at = list(depth_cm = 5))
em
```

\newpage

Let's work with the estimates and contrasts as two separate objects.

```{r}
em <- emmeans(mod2, specs = "trt_yr", type = "response") # estimates
pem <- pairs(em) # contrasts
```

The `summary` of a pairs object can provide CIs of the estimated differences when `infer = c(TRUE, TRUE)`:
(using `kable` to print only 3 digits - otherwise, output is too wide and p-vals end up on separate page)

```{r digits = 3}
knitr::kable(summary(pem, infer = c(TRUE, TRUE)), digits = 3) # using kable to fit on 1pg
```

\newpage

Now, it would be awesome to let emmeans do the work of estimating the difference between two curves at all of the values we care about. In fact, applying `emmeans` with `at = list(depth_cm = v)` for many values of `v` gives us the same difference curves from `smooth_diff`, the first method of inference we discussed.

```{r fig.height = 4, fig.width = 6}
vec <- seq(from = 0, to = 45, by = 2.5) # can use finer seq if desired

diffs_df <- purrr::map_dfr(vec, function(v) {
  em <- emmeans(mod2, specs = "trt_yr", type = "response", at = list(depth_cm = v))
  pem <- pairs(em)
  df <- as.data.frame(summary(pem, infer = c(TRUE, TRUE)))
  df$depth_cm <- v
  df
})

# grab only the rows that compare the treatments within the same year
str <- paste(var_df$var1, "-", var_df$var2)
diffs_df <- dplyr::filter(diffs_df, contrast %in% str)

ggplot(diffs_df) +
  geom_point(aes(x = depth_cm, y = estimate, color = contrast)) +
  geom_ribbon(aes(x = depth_cm, ymin = lower.CL, ymax = upper.CL), alpha = .5) +
  facet_wrap(~contrast) + geom_hline(yintercept = 0) +
  guides(color = FALSE) +
  ggtitle("Differences of expected sqrt(Resis_Mpa)")
```

\newpage

We can also use a similar strategy to extract the predictions and their CIs, and plot them on the same curve to visualize where differences are significant on the original scale.

```{r fig.height = 4, fig.width = 6}
vec <- seq(from = 0, to = 45, by = 2.5)

ests_df <- purrr::map_dfr(vec, function(v) {
  em <- emmeans(mod2, specs = "trt_yr", type = "response", at = list(depth_cm = v))
  df <- as.data.frame(em)
  df$depth_cm <- v
  
  # want trt information separately for faceeting later
  df <- tidyr::separate(df, col = trt_yr, sep = "y_", remove = FALSE,
                        into = c("trt", "yr_doy"))
  df
})

ests_df <- arrange(ests_df, depth_cm)

ggplot(ests_df, aes(group = trt, color = trt, fill = trt)) +
  geom_line(aes(x = depth_cm, y = response)) +
  geom_ribbon(aes(x = depth_cm, ymin = lower.CL, ymax = upper.CL),
              alpha = .5) +
  facet_wrap(~yr_doy)
```



```{r echo = FALSE}
# play a fanfare noise when everything is done running
beepr::beep(3)
```
